\textbf{Varying coefficients correlated velocity models in complex landscapes with application to narwhal behavioral response to ship exposure and airgun exposure}

### 2024-05-12 Latex/main.tex
latest version of the paper

The land geometry for the specific region in South-East Greenland is gathered from \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database. The land polygons define the boundaries of the domain $\mathcal{D}$ in which the narwhals can move. The closest point on the shoreline to the observed GPS position $y_{ij}$  is denoted $\hat{p}_{ij}$. Due to GPS measurement errors, it is an approximation of the unobserved closest point $p_{ij}$ to the true position, and therefore the distance to $\hat{p}_{ij}$, denoted $\hat{D}^{shore}_{ij}$ in the sequel, is considered a close approximation of the actual distance to the boundary $D^{shore}_{ij}=d(x_{ij},p_{ij})$. About $9\%$ of the GPS measurements in the dataset turned out to be on land. After visual examination, this was attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements. We therefore modified the land polygons using \texttt{qgis} based on a satellite image of Sentinel-2 downloaded from Google Earth Engine API. After modification,  $1.4 \%$ of the observations remained on land. Distance to shore values range from $0$ to $7.6$ km. We define an exposure to the shore covariate with the following formula
\begin{equation}
	\left\{
	\begin{array}{ll}
		E^{shore}_{ij}=0 & \mbox{if } D_{min} <D^{shore}_{ij}<D_{max} \\
		E^{shore}_{ij}=\frac{1}{D_{shore}(t_{ij})} & \mbox{if } D^{shore}_{ij} \leq D_{max}
	\end{array}
	\right.
	\label{eq: exp shore definition}
\end{equation}
where the reference thresholds are $D_{min}=58$ m, which is the $5\%$ quantile of observed distances, and $D_{max}=335$ m, which is the $25 \%$ quantile of observed distance. Thus we only consider the effect of the shore when the narwhals are close enough to it. We denote $\hat{E}^{shore}_{ij}$ the approximation of $E^{shore}_{ij}$ obtained by replacing the true distance $D^{shore}_{ij}$ by $\hat{D}^{shore}_{ij}$.
The angle between the narwhal's heading and the shoreline $\Theta_{ij}$ is defined as the angle between the vectors $x_{ij}-p_{ij}$ and the true velocity $v_{ij}$. It is estimated as the angle $\hat{\Theta}_{ij}$ between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_j=t_{i,j+1}-t_{ij}$, and the vector $n_{ij}=y_{ij}-\hat{p}_{ij}$. This is illustrated in figure~\ref{fig: illustration nearest shore points}. A value $\Theta_{ij}=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta_{ij} \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right[$ indicates movement toward the shore.

@@

\begin{equation*} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation bis}
\end{equation*}
where $X$ is the two-dimensional location and $V$ is the two-dimensional velocity.
In this section, we obtain explicit distributional results about the process $X$. Though this model is considered in several papers, we found no mention of the closed form formulas derived here. In \cite{gurarie_correlated_2017}, only the process $V$ is exhaustively studied, and the classical distributional results about this process are used for estimation and simulation purposes.
A more general formulation is considered by \cite{albertsen_generalizing_2018} in which the diffusion matrix is lower triangular matrix with positive diagonal elements whereas and two distinct autocorrelation parameters are considered in both directions allowing for anisotropic movement.
In this case, \cite{albertsen_generalizing_2018} derives an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}. But again, exact formulas for the process $X$ are not mentioned and the Euler-scheme is used to get approximate transition densities of $X$. Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\ref{eq: RACVM equation}, not only the velocity but also the position process has a closed form formula. \\

The transition density of the velocity Ornstein-Uhlenbeck process is  
\begin{equation}V(t+\Delta) \vert V(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
\end{equation}

@@

By integrating the velocity equation over time one gets the following conditional distribution for the location process :
\begin{equation}
	X(t+\Delta) \vert X(t)=x, V(t)=v \sim \mathcal{N}\left(x+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v-\mu),Q(\Delta)\right)  
\end{equation}
Here, the exact formula for the covariance matrix $Q(\Delta)$ is the following :

\[
Q(\Delta)=\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right) I_2
\]
The location process is not markovian, but the whole process $(X(t), V(t))^\top$ is.
The two components of the location and velocity processes are independent.
and one can check that for $\omega=0$, the formulas that are derived here match the ones of the continuous-time correlated random walk model \cite{johnson_continuoustime_2008}. Figure~\ref{fig: sample RACVM} shows two samples of a RCVM along with the histograms of the empirical velocity norm. Derivations of the formulas given in this paragraph and the next one are detailed in appendix~\ref{section: detailed RACVM computations}.\\

@@

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/illustrative_sample_standard.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/illustrative_sample_fjords.png}
		\caption{}
	\end{subfigure}
	\caption{(a) Samples of a RACVM with parameters $\tau=1 h$, $\nu=4$ km/h,$\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $h=5$ min. The red cross indicates the starting point of the process. The samples were obtained from the exact transition densities. (b) Samples of a CRCVM with parameters $\tau=1 h$, $\nu=4$ km/h,$\omega$ as a smooth function of covariates $\Theta$ and $D^{shore}$ as described in section~\ref{section: CRCVM}. Position are sampled from constant time step $h= 1$ min and then doownsampled to $5$ min.The samples were obtained from approximate transition densities  with constant covariates on each time step. }
	\label{fig: sample RACVM}
\end{figure}

@@

with 
\begin{equation} 
		A(X(t),V(t))=\begin{pmatrix} 
			\frac{1}{\tau} & -\omega(X(t),V(t)) \\
			\omega(X(t),V(t)) & \frac{1}{\tau}
		\end{pmatrix}
	\label{eq: CRCVM matrix A}
\end{equation}
where $\omega(X(t),V(t))=f_{\omega}(D^{shore}(X(t)),\Theta(X(t),V(t)))$, $D^{shore}(X(t))$ is the distance from $X(t)$ to the boundary of the domain $\partial \mathcal{D}$
and $\Theta(X(t),V(t))$ is the angle between the velocity and the shoreline as defined in subsection ~\ref{subsection: covariates}.Thus, $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.

@@

This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal. Any smooth function that meets minimum requirements could be used instead.An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions. Figure~\ref{fig: trueomega} shows a smooth function obtained from equation~\ref{eq: smooth omega} for the parameter $\omega$, while figure~\ref{fig: splineomega} represents a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions. In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM). A sample of a CRCVM with smooth function $f_{\omega}$ defined as by equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: sample RACVM}. The domain in which the process $X$ is constrained represents Scoresby-Sound fjords system coastline in South-East Greenland.

\begin{figure}[H]
	\centering
		\centering
		\includegraphics[scale=0.25]{images/simulation study/true_smooth_omega_rect.png}
		\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
		$D_0=0.3$ km, $\lambda=1.5$, $\kappa=0.5$ and $\omega_0=\pi$ rad/min}
		\label{fig: trueomega}
\end{figure}

\begin{figure}[H]
		\centering
		\includegraphics[scale=0.25]{images/simulation study/spline_smooth_omega_rect.png}
		\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
		This smooth function is constructed via bivariate tensor spline using a cubic spline basis with three degrees of freedom}
	\label{fig: splineomega}
\end{figure}

 The general model is very flexible, and different species-specific behaviours could be described. The function $\omega$ allows to control at which distance the animal will turn away from the shore, and how fast it will turn. We will show in section~\ref{section: simulation constrained motion} that is can be estimated from the observed positions, distances and angles. Specific constraints on the smooth function $f_{\omega}$ could also be set by biological expertise and knowledge about the species in question. Additionally, the tendency of an animal to move along the boundary of its spatial domain can be captured in the parameter $\tau$, as more persistent movement when the angle $\theta$ has value close to $\pm\frac{\pi}{2}$ would represent such an inclination. Again, this behavior can be statistically assessed from the observations. Examples of trajectories for different assumptions on the smooth parameters $\tau$ and $\nu$ are shown in Appendix.

@@

Denoting $n_{i,pre}$ the number of observations before exposure for the narwhal $i$, and $\tau_i$, $\nu_i$, $\omega_i$ the parameters of the diffusion .

\begin{equation}  \forall i \in \{1,\cdots,N\}, \quad \left\{
	\begin{array}{l}
		Y_{ij}=X_{ij}+\varepsilon_{ij} \quad j \in \{1,\cdots,n_{i,pre}\}\\
		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  \\
		dX_i(t)=V_i(t)dt  \\
		dV_i(t)=-\begin{pmatrix} 
			\frac{1}{\tau_i} & -\omega_i(t) \\
			\omega_i(t) & \frac{1}{\tau_i}
		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
		\log(\tau_i)=\log(\tau_{pre})+b_{\tau,pre}^{(i)} \\
		\log(\nu_i)=\log(\nu_{pre})+b_{\nu,pre}^{(i)}  \\
		\begin{pmatrix} b_{\tau,pre}^{(i)} \\ b_{\nu,pre}^{(i)} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau,pre}^2 & 0 \\ 0 & \sigma_{\nu,pre}^2 \end{pmatrix}\right) \\
		\omega_i(t)=f_{\omega}(E^{shore}_i(t),\hat{\Theta}_i(t))\\
		f_{\omega}(E^{shore},\Theta)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(E^{shore},\Theta), \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
	\end{array}
	\right.
	\label{eq: exact baseline model}
\end{equation}

$\log(\tau_{pre})$, $\log(\nu_{pre})$, $\omega_{0,pre}$ are population intercepts and $b^{(i)}_{\tau,pre}$, $b^{(i)}_{\nu,pre}$
are the individual parameters.

The functions $\psi_k$ in equation~\ref{eq: exact baseline model} are bivariate basis functions constructed through tensor products of univariate cubic spline basis functions. We refer to \cite{wood_generalized_2017} section $4.
1.8$ for the construction of tensor splines basis functions.

%Add plot of the spline basis functions \\
The splines degree of freedom is $L=q_E \times q_{\Theta}$ and the numbers of knots in the splines is $q_E-1$ and $q_{\Theta}-1$ for the respective covariates $E_{shore}$ and $\Theta$. It is needed to choose them before hand. One possibility to help choosing the best value is to increase the degree of freedom step by step  and find the value for which the decrease in AIC starts to get lower.\\

@@

\begin{equation}  \forall i \in \{1,\cdots,N\}, \quad \left\{
	\begin{array}{l}
		Y_{ij}=X_{ij}+\varepsilon_{ij} \quad j \in \{n_{i,pre}+1,\cdots,n_{i,post}\}\\
		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  \\
		dX_i(t)=V_i(t)dt  \\
		dV_i(t)=-\begin{pmatrix} 
			\frac{1}{\tau_i} & -\omega_i(t) \\
			\omega_i(t) & \frac{1}{\tau_i}
		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
		\log(\tau_{ij})=\log(\tau_{0,post}) +\sum_{k=1}^{q} \log(\tau_{k,post})\phi_{k}(E^{ship}_i(t))+b^{(i)}_{\tau,post} \\
		\log(\nu_{ij})=\log(\nu_{0,post}) +  \sum_{k=1}^{q} \log(\nu_{k,post}) \phi_{k}(E^{ship}_i(t)) +b^{(i)}_{\nu,post}  \\
		\omega_i(t)=f_{\omega}(E^{shore}_i(t),\Theta_i(t))\\
		f_{\omega}(E^{shore},\Theta)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(E^{shore}_i,\Theta_i), \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
	\end{array}
	\right.
	\label{eq: exact response model}
\end{equation}

Again, $\log(\tau_{0,post})$, $\log(\nu_{0,post})$, $\omega_{0,post}$ are population intercepts, $b^{(i)}_{\tau,post}$,$b^{(i)}_{\nu,post}$ are individual parameters. The functions $\phi_k$ are cubic spline basis functions and $q$ is the spline degree of freedom for the exposure to the ship covariate.

>>>>

The land geometry for the specific region in South-East Greenland is gathered from \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database. The land polygons define the boundaries of the domain $\mathcal{D}$ in which the narwhals can move. The closest point on the shoreline to the observed GPS position $y_{ij}$  is denoted $\hat{p}_{ij}$. Due to GPS measurement errors, it is an approximation of the unobserved closest point $p_{ij}$ to the true position, and therefore the distance to $\hat{p}_{ij}$, denoted $\hat{D}^{shore}_{ij}$ in the sequel, is considered a close approximation of the actual distance to the boundary $D^{shore}_{ij}=d(x_{ij},p_{ij})$. About $9\%$ of the GPS measurements in the dataset turned out to be on land. After visual examination, this was attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements. We therefore modified the land polygons using \texttt{qgis} based on a satellite image of Sentinel-2 with 10 metres resolution downloaded from Google Earth Engine API. After adjustements,  approximately $1 \%$ of the observations remained on land. Distance to shore values range from $0$ to $7.6$ km. We define an exposure to the shore covariate with the following formula
\begin{equation}
	\left\{
	\begin{array}{ll}
		E^{shore}_{ij}=0 & \mbox{if } D^{shore}_{ij}>D_{up} \\
		E^{shore}_{ij}=\frac{1}{D_{shore}(t_{ij})} & \mbox{if } D_{low}<D^{shore}_{ij} \leq D_{up} \\
		E^{shore}_{ij}=\frac{1}{D_{low}} & \mbox{ if } D^{shore}_{ij} \leq D_{low}
	\end{array}
	\right.
	\label{eq: exp shore definition}
\end{equation}
where the reference thresholds are $D_{low}=73$ m, which is the $5\%$ quantile of observed distances, and $D_{up}=830$ m, which is the median of observed distances. Thus we only consider the effect of the shore when the narwhals are close enough to it. We denote $\hat{E}^{shore}_{ij}$ the approximation of $E^{shore}_{ij}$ obtained by replacing the true distance $D^{shore}_{ij}$ by $\hat{D}^{shore}_{ij}$.
The angle between the narwhal's heading and the shoreline $\Theta_{ij}$ is defined as the angle between the vectors $x_{ij}-p_{ij}$ and the true velocity $v_{ij}$. It is estimated as the angle $\hat{\Theta}_{ij}$ between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_{ij}=t_{i,j+1}-t_{ij}$, and the vector $\vec{n}_{ij}=y_{ij}-\hat{p}_{ij}$. This is illustrated in figure~\ref{fig: illustration nearest shore points}. A value $\Theta_{ij}=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta_{ij} \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right[$ indicates movement toward the shore.

@@

\begin{equation*} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation bis}
\end{equation*}
where $X$ is the two-dimensional location and $V$ is the two-dimensional velocity.
In this section, we obtain explicit distributional results about the process $X$. Though this model is considered in several papers, we found no mention of the closed form formulas derived here. In \cite{gurarie_correlated_2017}, only the process $V$ is exhaustively studied, and the classical distributional results about this process are used for estimation and simulation purposes.
A more general formulation is considered by \cite{albertsen_generalizing_2018} in which the diffusion matrix is lower triangular matrix with positive diagonal elements whereas and two distinct autocorrelation parameters are considered in both directions allowing for anisotropic movement.
In this case, \cite{albertsen_generalizing_2018} derives an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}. But again, exact formulas for the process $X$ are not mentioned and the Euler-scheme is used to get approximate transition densities of $X$. Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\ref{eq: RACVM equation}, not only the velocity but also the position process have a closed form formula. \\

The transition density of the velocity Ornstein-Uhlenbeck process is  
\begin{equation}V(t+\Delta) \vert V(t)=v \sim \mathcal{N}\left( (I_2-\exp(-A\Delta))\mu + \exp(-A\Delta)v, \frac{2\nu^2}{\pi}(1-e^{-2\frac{\Delta}{\tau}}) I_2 \right) 
	\label{eq: transition RACVM velocity}
\end{equation}

@@

By integrating the velocity equation over time one gets the following conditional distribution for the location process :
\begin{equation}
	X(t+\Delta) \vert X(t)=x, V(t)=v \sim \mathcal{N}\left(x+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v-\mu),Q(\Delta)\right)  
	\label{eq: transition RACVM location}
\end{equation}
Here, the exact formula for the covariance matrix $Q(\Delta)$ is the following :

\[
Q(\Delta)=\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right) I_2
\]
The location process is not markovian, but the whole process $\begin{pmatrix}X(t) &V(t) \end{pmatrix}^\top$ is.
The two components of the location and velocity processes are independent, and one can check that for $\omega=0$, the formulas that are derived here match the ones of the continuous-time correlated random walk model \cite{johnson_continuoustime_2008}. Figure~\ref{fig: sample RACVM} shows one sample of a RCVM along with a sample of a constrained version of a RCVM that is defined in section~\ref{section: CRCVM}. Derivations of the formulas given in this paragraph and the next one are detailed in appendix~\ref{section: detailed RACVM computations}.\\

@@

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/rcvm/illustrative_sample_standard.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/crcvm/illustrative_sample_fjords_standard.png}
		\caption{}
	\end{subfigure}
	\caption{(a) Samples of a RACVM with parameters $\tau=1 h$, $\nu=4$ km/h,$\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $h=5$ min. The red cross indicates the starting point of the process. The samples were obtained from the exact transition densities in equations~\ref{eq: transition RACVM velocity} and~\ref{eq: transition RACVM location}. (b) Samples of a constrained RCVM with parameters $\tau=1 h$, $\nu=4$ km/h, and $\omega$ as a smooth function of covariates $\Theta$ and $D^{shore}$ as described in section~\ref{section: CRCVM}. Position are sampled from constant time step $h= 1$ min and then downsampled to $5$ min. The samples were obtained from approximate transition densities  with constant covariates on each time step.}
	\label{fig: sample RACVM}
\end{figure}

@@

with 
\begin{equation} 
		A(X(t),V(t))=\begin{pmatrix} 
			\frac{1}{\tau} & -\omega(X(t),V(t)) \\
			\omega(X(t),V(t)) & \frac{1}{\tau}
		\end{pmatrix}
	\label{eq: CRCVM matrix A}
\end{equation}
where $\omega(X(t),V(t))=f_{\omega}(D^{shore}(t),\Theta(t))$, $D^{shore}(t)$ is the distance from $X(t)$ to the boundary of the domain $\partial \mathcal{D}$
and $\Theta(t)$ is the angle between the velocity $V(t)$ and the shoreline as defined in subsection ~\ref{subsection: covariates}. Thus, $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.

@@

This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal. Any smooth function that meets minimum requirements could be used instead. An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions. Figures~\ref{fig: smoothomega1} and~\ref{fig: smoothomega2} show a smooth function obtained from equation~\ref{eq: smooth omega}, while figures~\ref{fig: smoothomegasplines1} and~\ref{fig: smoothomegasplines2}  represents a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions. In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ meeting minimum requirements will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM). A sample of a CRCVM with smooth function $f_{\omega}$ defined by equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: sample RACVM}. The domain in which the process $X$ is constrained represents Scoresby-Sound fjords system coastline in South-East Greenland.

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/crcvm/smooth_omega1.png}
		\caption{}
		\label{fig: smoothomega1}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/crcvm/smooth_omega2.png}
		\caption{}
	\label{fig: smoothomega2}
	\end{subfigure}

	\begin{subfigure}{0.48\textwidth}
	\centering
	\includegraphics[scale=0.3]{images/crcvm/smooth_splines_omega1.png}
	\caption{}
	\label{fig: smoothomegasplines1}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
	\centering
	\includegraphics[scale=0.3]{images/crcvm/smooth_splines_omega2.png}
	\caption{}
	\label{fig: smoothomegasplines2}
	\end{subfigure}
	\caption{Examples of smooth functions $f_{\omega}$. Angular velocity $\omega$ increases as $\Theta$ approaches $\pm\pi$, and decreases with increasing distance to the shore. (a)-(b) Two views of the smooth function $f_{\omega}$ defined by equation~\ref{eq: smooth omega}. (c)-(d) Two views of a smooth function $f_{\omega}$ obtained by combining nine bivartiate cubic spline basis functions.}

\end{figure}

 This general model is very flexible, and different species-specific behaviours could be described. The function $\omega$ allows to control at which distance the animal will turn away from the shore, and how fast it will turn. We will show in section~\ref{section: simulation constrained motion} that is can be estimated from the observed positions, observed distances and angles. Specific constraints on the smooth function $f_{\omega}$ could also be set by biological expertise and knowledge about the species in question. Additionally, the tendency of an animal to move along the boundary of its spatial domain can be captured in the parameter $\tau$, as more persistent movement when the angle $\theta$ has value close to $\pm\frac{\pi}{2}$ would represent such an inclination. Again, this behavior can be statistically assessed from the observations.

 More precisely, to model persistent movement along the shore line, we modify equation~\ref{eq: CRCVM matrix A} by expressing $\tau$ as a smooth function :
 \[\tau(X(t),V(t))=f_{\tau}(\Theta(t))\] where $f_{\tau}$ is a smooth function from $[-\pi,\pi]$ to $\R$.
 The function $f_{\tau}$ is defined such that it reaches local maximas in $\Theta=\pm\frac{\pi}{2}$, which represents movement along the boundary of the domain. An example of such a function is given in figure~\ref{fig:smoothtaubump}.
 \begin{figure}[H]
 	\centering
 	\includegraphics[scale=0.3]{images/crcvm/smooth_tau_bump}
 	\caption{Example of smooth function $f_{\tau}$ that can represent persistent movement along the boundary of the domain.}
 	\label{fig:smoothtaubump}
 \end{figure}

 We show side by side the samples obtained within different domain for a constant value of $\tau$ and for a smooth function $\tau$ defined as in figure~\ref{fig:smoothtaubump}.

 \begin{figure}[H]
 	\centering
 	\begin{subfigure}{0.48\textwidth}
 		\centering
 		\includegraphics[scale=0.35]{images/crcvm/illustrative_sample_circ_standard}
 		\label{fig:illustrativesamplecircstandard}
 	\end{subfigure}
 	\begin{subfigure}{0.48\textwidth}
 		\centering
 		\includegraphics[scale=0.35]{images/crcvm/illustrative_sample_circ_pers}
 		\label{fig:illustrativesamplecircpers}
 	\end{subfigure}

 	\begin{subfigure}{0.48\textwidth}
 		\centering
 		\includegraphics[scale=0.35]{images/crcvm/illustrative_sample_rect_standard}
 		\label{fig:illustrativesamplerectstandard}
 	\end{subfigure}
 	\begin{subfigure}{0.48\textwidth}
 		\centering
 		\includegraphics[scale=0.35]{images/crcvm/illustrative_sample_rect_pers}
 		\label{fig:illustrativesamplerectpers}
 	\end{subfigure}

 	\begin{subfigure}{0.48\textwidth}
 		\centering
 		\includegraphics[scale=0.3]{images/crcvm/illustrative_sample_fjords_standard}
 		\label{fig:illustrativesamplefjordsstandard}
 	\end{subfigure}	
 	\begin{subfigure}{0.48\textwidth}
 		\centering
 		\includegraphics[scale=0.3]{images/crcvm/illustrative_sample_fjords_pers}
 		\label{fig:illustrativesamplefjordspers}
 	\end{subfigure}
 	\caption{Left: samples from a CRCVM with smooth $\omega$ and constant $\tau=1$ h in three different domains. Right: samples from a CRCVM with smooth $\omega$ and smooth $\tau$ to represent persistent movement along the shoreline.}
 	\label{fig: crcvm examples}
 \end{figure}

@@

Denoting $n_{i,pre}$ the number of observations before exposure for the narwhal $i$, and $\tau_i$, $\nu_i$, $\omega_i$ the parameters of the diffusion. The complete statistical model for the baseline estimations is the following:

\begin{equation}  \forall i \in \{1,\cdots,N\}, \quad \left\{
	\begin{array}{l}
		Y_{ij}=X_{ij}+\varepsilon_{ij} \quad j \in \{1,\cdots,n_{i,pre}\}\\
		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  \\
		dX_i(t)=V_i(t)dt  \\
		dV_i(t)=-\begin{pmatrix} 
			\frac{1}{\tau_i} & -\omega_i(t) \\
			\omega_i(t) & \frac{1}{\tau_i}
		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
		\log(\tau_i)=\log(\tau_0)+b_{\tau,pre}^{(i)} \\
		\log(\nu_i)=\log(\nu_0)+b_{\nu,pre}^{(i)}  \\
		\begin{pmatrix} b_{\tau,pre}^{(i)} \\ b_{\nu,pre}^{(i)} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau,pre}^2 & 0 \\ 0 & \sigma_{\nu,pre}^2 \end{pmatrix}\right) \\
		\omega_i(t)=f_{\omega}(\hat{E}^{shore}_i(t),\hat{\Theta}_i(t)) \quad t \geq 0\\
		f_{\omega}(E^{shore},\Theta)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(E^{shore},\Theta), \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
	\end{array}
	\right.
	\label{eq: exact baseline model}
\end{equation}

The coefficients $\log(\tau_{0})$, $\log(\nu_{0})$, $\omega_{0}$ are population intercepts and $b^{(i)}_{\tau,pre}$, $b^{(i)}_{\nu,pre}$
are the individual parameters.

The functions $\psi_k$ in equation~\ref{eq: exact baseline model} are bivariate basis functions constructed through tensor products of univariate cubic spline basis functions. We refer to \cite{wood_generalized_2017} section $4.
1.8$ for the construction of tensor splines basis functions. The splines degree of freedom is $L=q_E \times q_{\Theta}$ and the numbers of knots in the splines is $q_E-1$ and $q_{\Theta}-1$ for the respective covariates $E_{shore}$ and $\Theta$. It is needed to choose them before hand.\\

@@

\begin{equation}  \forall i \in \{1,\cdots,N\}, \quad \left\{
	\begin{array}{l}
		Y_{ij}=X_{ij}+\varepsilon_{ij} \quad j \in \{n_{i,pre}+1,\cdots,n_{i,post}\}\\
		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  \\
		dX_i(t)=V_i(t)dt  \\
		dV_i(t)=-\begin{pmatrix} 
			\frac{1}{\tau_i} & -\omega_i(t) \\
			\omega_i(t) & \frac{1}{\tau_i}
		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
		\log(\tau_{ij})=\log(\tau_{0}) +\sum_{k=1}^{q} \log(\tau_{k,post})\phi_{k}(E^{ship}_i(t))+b^{(i)}_{\tau,post} \\
		\log(\nu_{ij})=\log(\nu_{0}) +  \sum_{k=1}^{q} \log(\nu_{k,post}) \phi_{k}(E^{ship}_i(t)) +b^{(i)}_{\nu,post}  \\
		\omega_i(t)=f_{\omega}(\hat{E}^{shore}_i(t),\hat{\Theta}_i(t)) \quad t \geq 0\\
		f_{\omega}(E^{shore},\Theta)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(E^{shore}_i,\Theta_i), \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
	\end{array}
	\right.
	\label{eq: exact response model}
\end{equation}

Again, $b^{(i)}_{\tau,post}$,$b^{(i)}_{\nu,post}$ are individual parameters. The functions $\phi_k$ are cubic spline basis functions and $q$ is the spline degree of freedom for the exposure to the ship covariate.

\subsection{State-space model}
\label{section: state space model}
In most applications, it is needed to consider measurement errors through a state-space model as done by \cite{johnson_continuoustime_2008}. This is the approach we use here. In the statistical model~\ref{eq: exact baseline model} for the baseline model, we make the approximation mentioned in \cite{michelot_varying-coefficient_2021} that the parameter $\omega_i$ is constant on each time step $[t_{ij},t_{i,j+1}]$, equal to its value at time $t_{ij}$, that is 
\[\forall t \in [t_{ij},t_{i,j+1}], \quad \omega_{i}(t) \simeq \omega(t_{ij})= f_{\omega}(\hat{E}^{shore}_{ij},\hat{\Theta}_{ij})\]
 Denote $\Delta_{ij}=t_{i,j+1}-t_{ij}$, $V_{ij}=V(t_{ij})$, $X_{ij}=X(t_{ij})$, ,$\omega_{ij}=\omega(t_{ij})$, $A_{ij}=\begin{pmatrix} 
	\frac{1}{\tau_{i}} & -\omega_{ij} \\
	\omega_{ij} & \frac{1}{\tau_{i}}
\end{pmatrix}$ and $\alpha_{ij}=\begin{pmatrix} X_{ij,1}  & X_{ij,2} & V_{ij,1} & V_{ij,2}\end{pmatrix}^\top$ where $\log(\tau_i)=\log(\tau_0)+b_{\tau,pre}^{(i)}$ and $\log(\nu_i)=\log(\nu_0)+b_{\nu,pre}^{(i)}$.
\begin{comment}
\begin{equation}
	\forall j \in \{1, \cdots, n_{i,pre}\}, \quad 
	\left\{ \begin{array}{l}
		V_{i,j+1}=\exp(-A_{ij}\Delta_{ij})V_{ij}+\zeta_{ij} \\
		X_{i,j+1}=X_{ij}+A_{ij}^{-1}(I_2-\exp(-A_{ij}\Delta_{ij}))V_{ij}+\xi_{ij}
	\end{array}\right. \\
	\label{eq: RACVM autoregressive formula}
\end{equation}
